A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer

### 2024-01-25 content/manuscript.qmd
work in Jan's edits

Head and neck squamous cell carcinomas (HNSCC) frequently spread through the lymphatic system [@lindberg_distribution_1972;@woolgar_histological_1999]. Current diagnostic imaging modalities are unable to detect microscopic nodal metastases, which requires pathological examination of extracted tissue [@snyder_petct_2021;@strohl_petct_2021]. Since the recurrence of nodal disease is detrimental to a patient's outcome [@ho_decision_2014], large volumes in the head and neck region are often irradiated electively to minimize the risk of missing occult disease. Decision guidelines about which nodal regions -- i.e., anatomically defined lymph node levels (LNLs) -- to irradiate [@biau_selection_2019] are currently not based on a patient's individual risk, but only on the overall prevalence of nodal disease as reported in the literature [@lindberg_distribution_1972;@woolgar_histological_1999].

To personalize this prediction of the risk for occult diease, given a patient's individual diagnosis, we published

@@

Such a personalized risk prediction may allow clinitians to safely reduce the elective clinical target volume (CTV-N) and thus reduce side-effects that degrade the patient's quality of life [@batth_practical_2014].

HNSCC patients with primary tumors at different subsites, e.g. in the oral cavity and in the oropharynx, also show different patterns of lymphatic spread [@lindberg_distribution_1972;@woolgar_histological_1999]. Our model does so far not have the capability to naturally describe different tumor subsites. To that end, we present an approach using mixtures of these HMMs in this work. This makes intuitive sense, because if a tumor lies anatomically between two anatomically close subsites with slightly different spread patterns, we may be able to describe its lymphatic progression as a mixture of these two spread patterns.

@@

To compute the personalized risk of occult disease $\mathbf{X}$, given a diagnosis $\mathbf{Y}$, we can begin by stating Bayes' law:

@@

In the above equation, the term $P \left( \mathbf{Y} \mid \mathbf{X} \right)$ is given by the sensitivity and specificity of the diagnosis. The crucial term that our model attempts to compute, is the prior probability of involvement $P \left( \mathbf{X} \right)$ (or rather of the hidden state of involvement, see the [next section](#sec-hmm)).

@@

Each LNL $v \in \{ 1, 2, \ldots, V\}$ considered in our model is represented by a binary random variable (RV) $X_v[t]$ taking on the true state of that level at the abstract time-step $t$ (0 for "healthy" and 1 for "involved"). Collected in a random vector $\mathbf{X}[t] = \left( X_1[t], X_2[t], \ldots, X_V[t] \right)$ they form the patient's state w.r.t. their lymphatic involvement at time $t$.

We model the process of tumor progression via lymphatic drainage by connecting the RVs in a graph, as shown in @fig-schematic-and-graph. The arcs in [this graph](#fig-schematic-and-graph) represent conditional probabilities. The orange arcs correspond to observing a diagnosis $Y_v$, given the true state $X_v$. For the sake of brevity, we will not go into the details of describing how to infer the true -- but technically hidden -- state of LNL involvement from diagnoses with lower-than-one sensitivity and specificity. This description can be found in @zora231470. Throughout this work, we instead combine diagnostic and pathologic involvement information from the data into a "maximum likelihood" diagnosis and assume its sensitivity and specificity to be one, meaning the normally hidden state $X_v$ becomes the observed state. This simplification is reasonable for pathologic involvement and because we are at this stage more interested in testing whether our model is able to describe a realistic probability distribution over lymphatic involvement.

The red arcs in the graph of @fig-schematic-and-graph depict the probability that the primary tumor spreads within one abstract time-step. While the blue arcs symbolize the spread from an upstream LNL -- given it is already metastatic -- to a downstream level. For example, the edge from $X_2$ to $X_3$ encodes the probability $P\left( X_3[t+1] \mid X_2[t] \right)$, which is parametrized with $b_3$, and $t_{2 \rightarrow 3}$ and tabulated in @tbl-spread-2-to-3. There is an additional restriction on any LNL's state $X_v[t+1]=0$ to be healthy: It requires that the level was also healthy in the time-step before, meaning $X_v[t]=0$. This is because we assume no spontaneous self-healing of metastatic levels.

+--------------+------------+-----------------------------------------------------+
|              | $X_2[t]=0$ | $X_2[t]=1$                                          |
+==============+:==========:+:===================================================:+
| $X_3[t+1]=0$ | $1-b_3$    | $(1-b_3)(1-t_{2 \rightarrow 3})$                    |
+--------------+------------+-----------------------------------------------------+
| $X_3[t+1]=1$ | $b_3$      | $1-b_3-t_{2 \rightarrow 3}+b_3 t_{2 \rightarrow 3}$ |
+--------------+------------+-----------------------------------------------------+

: Conditional probability $P\left( X_3[t+1] \mid X_2[t] \right)$ for a spread from LNL II to III during the transition from $t$ to $t+1$. This corresponds to one of the blue arcs in @fig-schematic-and-graph. Note that the values in the row with $X_3[t+1]=0$ is all zeros, and the row with $X_3[t+1]=1$ all ones if $X_3[t]=1$. {#tbl-spread-2-to-3 tbl-colwidths="[25,18,40]"}

With the introduced conditional probabilities, we can now compute the joint probability of any complete state $\mathbf{X}[t]=\boldsymbol{\xi}_i$ transitioning to any other possible state $\mathbf{X}[t+1]=\boldsymbol{\xi}_j$. Here, when we use $\boldsymbol{\xi}$ instead of $\mathbf{x}$ for the values the random vector $\mathbf{X}$ can take on, the $i$ and $j$ enumerate all $2^V$ combinations of the $V$ binary RVs. In the graph shown in @fig-schematic-and-graph, this amounts to $2^V=8$ distinct $\boldsymbol{\xi}_i$. Because these terms are essentially products of terms like those in @tbl-spread-2-to-3. We can then collect these terms in a _transition matrix_ $\mathbf{A}$:

@@

Note that this matrix still depends on the $b_v$ and $t_{r \rightarrow v}$ parameters, although we have dropped the explicit dependcy to keep the equations brief. Now, assuming that every patient started their disease with all LNLs being healthy, we can define the _starting distribution_ $\boldsymbol{\pi}$:

@@

And set every entry of this starting distribution to zero, except the first one, which we set to one. This means at $t=0$ there is a probability of one to be in the completly healthy state $\boldsymbol{\xi}_0 = \begin{pmatrix} 0, 0, \ldots, 0 \end{pmatrix}$.

Using the quantities introduced so far, the probability distribution vector with elements $P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right)$ after $t$ time-steps can now be conveniently expressed as a matrix product:

@@

This evolution implicitly marginalizes over all possible paths to arrive at state $\boldsymbol{\xi}_i$ after $t$ time-steps. Additionally, we also need to marginalize over the time of diagnosis -- which is unknown -- using a time-prior $P_T(t)$. Fortunately, the exact length and shape of this distribution on its own has little impact. But because we assume that early and advanced T-category patients are fundamentally the same, just on average diagnosed at different times $t$, we use the time-prior only to separate the respective patient's evolutions:

$$
P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid T, \boldsymbol{\theta} \right) = \sum_{t=0}^{t_\text{max}} P_T(t) \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
$$ {#eq-marginalized-evolution}

where $T \in \{ \text{early}, \text{advanced} \}$ denotes the T-category. We fix $P_\text{early}(t)$ to a binomial distribution with parameters $n=t_\text{max}=$ {{< var model.max_t >}} and $p_\text{early}=$ {{< var model.first_binom_prob >}}, while the advanced T-category's time-prior is also a binomial distribution where the $p_\text{advanced}$ parameter is learned.

Note that @eq-marginalized-evolution still depends on the parametrization of the transition matrix. We collect these parameters in a vector $\boldsymbol{\theta}=\{ b_v, t_{r \rightarrow v}, p_\text{advanced} \}$.

To train our model, we need to compute the likelihood of a dataset $\mathbf{D}=\left( \mathbf{x}_1, \mathbf{x}_2, \ldots, \mathbf{x}_N \right)$ of $N$ patients, given the model's parameters $\boldsymbol{\theta}$. This is given by a product of the terms in @eq-marginalized-evolution:

$$
P \left(\mathbf{D} \mid \boldsymbol{\theta} \right) = \prod_{i=1}^N P \left( \mathbf{X} = \mathbf{x}_i \mid T_i, \boldsymbol{\theta} \right)
$$ {#eq-likelihood}

Typically, one would compute the logarithm of this quantity for computational reasons. There is a wide array of available methods to obtain maximum likelihood estimates from this function or sample from the posterior over $\boldsymbol{\theta}$. In this work, we use Markov chain Monte Carlo sampling (MCMC) via the `emcee` Python package [@foreman-mackey_emcee_2013].

@@

@@

Here, the $\mathbf{c} \in [0,1]^g$ is the vector of mixing proportions with $\sum_{j=0}^g c_j = 1$, while the $\boldsymbol{\Psi} = \left( \boldsymbol{\theta}_1, \ldots, \boldsymbol{\theta}_g\right)$ is the vector of all $g$ models' parameters.

>>>>

Head and neck squamous cell carcinomas (HNSCC) frequently spread through the lymphatic system [@lindberg_distribution_1972;@woolgar_histological_1999]. Current diagnostic imaging modalities are unable to detect microscopic lymph node metastases [@snyder_petct_2021;@strohl_petct_2021]. To avoid nodal recurrences [@ho_decision_2014], large volumes in the neck are irradiated electively, which are at risk of harbouring occult disease. Guidelines about which lymph node levels (LNLs) to irradiate [@biau_selection_2019] are currently not based on a patient's individual risk, but only on the overall prevalence of nodal disease as reported in the literature [@lindberg_distribution_1972;@woolgar_histological_1999].

To personalize this prediction of the risk for occult disease, given a patient's individual diagnosis, we published

@@

Such a personalized risk prediction may allow clinicians to safely reduce the elective clinical target volume (CTV-N) and thus reduce side-effects that degrade the patient's quality of life [@batth_practical_2014].

Here, we extend the previous work by incorporating the primary tumor location (specified as ICD-10 code) into the model of lymphatic tumour progression, focusing on tumours in the oropharynx and the oral cavity. HNSCC patients with primary tumors at different subsites show different patterns of lymphatic spread [@lindberg_distribution_1972;@woolgar_histological_1999]. So far, this could be handled by training different models for broader categories of tumour locations, e.g. oropharynx and oral cavity tumours. However, this approach does not describe differences in lymphatic spread between different subsites within the oropharynx and oral cavity. To address this issue, we present an approach using mixtures of HMMs. The intuition is that the lymphatic spread of a tumor that lies anatomically at the boarder of oropharynx and oral cavity (e.g. tumours in the palate) may be described by a mixture of different models.

@@

Each LNL $v \in \{ 1, 2, \ldots, V\}$ considered in our model is represented by a binary random variable (RV) $X_v$ representing the true state of that level (0 for "healthy" and 1 for "involved"). A patient's state of lymph node involvement can be represented in a random vector $\mathbf{X} = \left( X_1, X_2, \ldots, X_V \right)$. When a patient is diagnosed with HNSCC, we only observe the clinical lymph node involvement based on imaging, which we denote as another binary random variable $Y_v$. To compute the personalized risk of occult disease $\mathbf{X}$, given a diagnosis $\mathbf{Y}$, we apply Bayes' law:

@@

In the above equation, the term $P \left( \mathbf{Y} \mid \mathbf{X} \right)$ is given by the sensitivity and specificity of the diagnostic procedure. The term $P \left( \mathbf{X} \right)$ represents the prior probability of involvement, which depends on the probability of the tumour to spread through the lymphatic system. The main task of the HMM is to model $P \left( \mathbf{X} \right)$ and the main contribution of this paper is to incorporate the primary tutor subset into the model of $P \left( \mathbf{X} \right)$.

@@

A patient's state of lymph node involvement $\mathbf{X}[t]$ evolves over discrete time steps $t$. Let us enumerate all $2^V$ possible states, representing all combinations of LNLs. In this paper, we consider ipsilateral LNLs I, II, III and IV, which amounts to 16 possible states. The HMM is specified by _transition matrix_ $\mathbf{A}$:

@@

which contains the conditional probabilities that a state $\mathbf{X}[t]=\boldsymbol{\xi}_i$ transitions to $\mathbf{X}[t+1]=\boldsymbol{\xi}_j$ over one time step. The transition matrix is specified and parameterised via the graphical model shown in @fig-schematic-and-graph. The red arcs in the graph of @fig-schematic-and-graph are associated the probability that the primary tumor spreads directly to a LNL (parameters $b_v$). The blue arcs describe the spread from an upstream LNL -- given it is already metastatic -- to a downstream level (parameters $t_{v-1 \rightarrow v}$).

Now, let $\boldsymbol{\pi}$ be the _starting distribution_

@@

denoting the probability to start in state $\boldsymbol{\xi}_i$ at time step 0. Assuming that every patient started their disease with all LNLs being healthy, we set $\pi_i$ to zero for all states  except the completly healthy state $\boldsymbol{\xi} = \begin{pmatrix} 0, 0, 0, 0 \end{pmatrix}$, which has probability 1.

Using the quantities introduced so far, the probability $P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right)$ to be in state $i$ in time step $t$ can now be conveniently expressed as a matrix product:

@@

This evolution implicitly marginalizes over all possible paths to arrive at state $\boldsymbol{\xi}_i$ after $t$ time-steps. Additionally, we must marginalize over the unknown time of diagnosis using a time-prior $P(t)$. This finally defines the probability distribution over all states of lymph node involvement used in @eq-bayes-law.

$$
P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta} \right) = \sum_{t=0}^{t_\text{max}} P_T(t) \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
$$ {#eq-marginalized-evolution}

where $\boldsymbol{\theta}=\{ b_v, t_{r \rightarrow v} \}$ denotes the set of all model parameters (7 in our case). Fortunately, the exact length and shape of this distribution on its own has little impact as previously shown. We set $t_\text{max}=$ {{< var model.max_t >}} and $P_\text{early}(t)$ to a binomial distribution with parameter {{< var model.first_binom_prob >}}. Further details on the HMM can be found in ...

@@

Let us now assume that primary tumors at different subsites have different patterns of lymphatic spread, corresponding to different model parameters $\boldsymbol{\theta}$. Training a separate model for every possible subsite (ICD-10 code) would require a sufficiently large dataset for every tumor site. However, anatomically nearby locations are expected to show very similar patterns of LNL involvement. Therefore, we consider a mixture model.

Let us assume that we have a dataset $\mathbf{D}$ that is specified via the number of patients $N_{is}$ that were diagnosed in LNL involvement state $i$ and had a primary tutor in subsite $s$. Let us further assume that we want to describe this dataset using a mixture of $M$ HMMs, each with a different set of model parameters $\boldsymbol{\theta_m}$. As the generative model of the data, we assume that a patient with subsite $s$ is generated with probability $\pi_{sm}$ from model $m$. The likelihood of the dataset can then be written as

$$
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_s \prod_i \left[ \sum_{m=1}^M \pi_{sm} P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m \right) \right]^{N_{is}}
$$ {#eq-mixture-distribution}

We now have two types of parameters, the probabilities of tutor spread for the different models $\boldsymbol{\theta_m}$, and the mixing coefficients $\pi_{sm}$. Assuming a uniform prior in the interval $[0,1]$ for all parameters, the posterior distribution over the parameters $P \left( \boldsymbol{\theta}, \boldsymbol{\pi} \mid \mathbf{D} \right)$ is given by the likelihood @eq-mixture-distribution except for a normalisation constant. In this work, we use Markov chain Monte Carlo sampling (MCMC) via the `emcee` Python package [@foreman-mackey_emcee_2013] to sample model parameters from the posterior distribution. However, $P \left( \boldsymbol{\theta}, \boldsymbol{\pi} \mid \mathbf{D} \right)$ itself is a multi-model distribution because one can permute the different models. To address this problem, we revert to an _expectation-maximization (EM)_ algorithm where we iteratively sample model parameters $\boldsymbol{\theta_m}$ using MCMC and then determine the most likely mixing coefficients.

...

@@

Here, the $\mathbf{c} \in [0,1]^g$ is the vector of mixing proportions with $\sum_{j=0}^g c_j = 1$, while the $\boldsymbol{\Psi} = \left( \boldsymbol{\theta}_1, \ldots, \boldsymbol{\theta}_g\right)$ is the vector of all $g$ models' parameters. Note that we will implement our model such that some of the parameters in each $\boldsymbol{\theta}_j$ are shared across the $g$ components -- namely the $t_{r \rightarrow v}$ corresponding to the blue arcs in @fig-schematic-and-graph.
